---
title: "Stat-Question-2"
format: html
editor: visual
---

The purpose of this file is to run through the statistical tests to answer the second proposed question about the dataset for my capstone project: Is there a relationship between an MS patient's status as a current or former smoker and their likelihood of experiencing the COVID-19 symptom of loss of taste or smell

#### Set up

```{r}
#Clear the environment
rm(list = ls())
#Load in packages 
library(tidyverse)
library(here)
library(ggfortify)
```

#### Load in the cleaned dataset

Steps 1 and 2 of the course's data analysis workflow (ensuring the data is entered correctly and constructing visualizations of the data, repairing data as needed) were completed prior and saved as the dataset that will be loaded in.

```{r}
covid <- read.csv(here("Data","MS_COVID_DATASET_FINAL.csv"))
```

#### Plot relationships and make guessess

```{r}
#Gather counts of confirmed loss of taste or smell symptoms (Yes, No, or empty), by current or former smoker status (Yes or Unanswered)
counts_smoker_status <- covid |> 
  group_by(current_or_former_smoker, covid19_sympt_loss_smell_taste) |>
    summarize(
      count = length(covid19_sympt_loss_smell_taste)
    )
#Create a bar graph to view counts of confirmed lost of taste or smell symptoms (Yes, No, or empty) by current or former smoker status (Yes or Unanswered)
ggplot(counts_smoker_status, aes(x = current_or_former_smoker, y = count, fill = covid19_sympt_loss_smell_taste)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  scale_fill_manual(values = c("yes" = "green", 
                               "no" = "red", 
                               "Unanswered" = "grey")) + 
  geom_text(aes(label = count), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5) +
  theme_bw()
```

Examining the graph, if we exclude those who left their symptom status unanswered, there is a good distribution between current or former smokers (93 points) and those who left their smoker status unanswered (130 data points).

Question: I think it would be best to exclude the unanswered symptom responses from each group (in particular, an unanswered symptom response among those who left their current or former smoker status unanswered means nothing). Once again, I don't think this is something we covered in class, so I need to seek help or online resources to accomplish this.

For now, I will carry out a chi-square goodness of fit test for the data displayed in the above diagram since we have categorical predictor and response variables.

Question: If we remove the unanswered symptom responses, would it then be appropriate to conduct a logistic regression? Or is a chi square goodness of fit test still the best approach?

#### Conduct a chi-square goodness of fit test

```{r}
# Transform the dataframe into a matrix
covid.mat <- xtabs(count ~ current_or_former_smoker + covid19_sympt_loss_smell_taste,
                   data = counts_smoker_status)
# Run the chi-square test
chisq.test(covid.mat)
#Access the specifics of the chi-squared test
covid.chi <- chisq.test(covid.mat)
names(covid.chi)
```

Notice, we have enough data points for each category that the chi square goodness of fit test is reliable. However, as previously mentioned, we should remove the unanswered symptom responses to better understand the relationship between experiencing the symptom of loss of taste or smell and an MS patient's status as a current or former smoker. Based on this chi square test, there is no significant difference between an MS patient's current or former smoker status and their likelihood of experiencing the COVID-19 symptom of loss of taste or smell (X-squared = 1.787, df = 2, p-value = 0.4092).

#### Construct a clean plot to convey these results

```{r}
#Create bar plot similar to the initial plot but with a few tweaks to make it more visually appealing
ggplot(counts_smoker_status, aes(x = current_or_former_smoker, y = count, fill = covid19_sympt_loss_smell_taste)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  scale_fill_manual(values = c("yes" = "green", 
                               "no" = "red", 
                               "Unanswered" = "grey")) + 
  geom_text(aes(label = count), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5) +
  xlab("Current or Former Smoker Status") +
  ylab("Number of Patients") +
  scale_x_discrete(limits = c("Unanswered", "yes"),
  labels = c("No Response", "Yes")) +
  theme(
    legend.key = element_rect(fill = NA),
    axis.title.x = element_text(size = 10), 
    axis.title.y = element_text(size = 10), 
    axis.text.x = element_text(colour = "Black"),
    panel.background = element_rect(fill = NA, colour = "black"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "gray")
    )
```

\`\`\`
